Viscous Transport in Eroding Porous Media

### 2019-12-27 docs/transport/erosion_transport.tex
Responses for JFM paper

%\documentclass[preprint,10pt]{elsarticle}
\documentclass{jfm}

@@

%  Alan N. Other\aff{1}
%  \corresp{\email{jfm@damtp.cam.ac.uk}},
%  H. - C. Smith\aff{1}
% \and J. Q.  Public\aff{2}}

@@

%\title{Viscous Transport in Eroding Porous Media}
%
%\author[SH]{Shang-Huan Chiu}
%\author[Nick]{M.~N.~J.~Moore}
%\author[Bryan]{Bryan D.~Quaife}
%
%\address[SH]{Department of Scientific Computing, Florida State
%University, Tallahassee, FL, 32306.}
%\address[Nick]{Department of Mathematics and Geophysical Fluid Dynamics
%Institute, Florida State University, Tallahassee, FL, 32306.}
%\address[Bryan]{Department of Scientific Computing and Geophysical Fluid
%Dynamics Institute, Florida State University, Tallahassee, FL, 32306.}

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared with models available in
  the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.

%  We extend our previous work, [{\em Journal of Computational Physics},
%  {\bf 375}:1--21, 2018], to simulate fluid-mechanical erosion of
%  two-dimensional porous media with nearly-touching grains.  The method
%  continues to use a boundary integral equation (BIE) method to solve
%  the fluid equations, and a stable second-order time stepping method to
%  erode the grains.  In our previous work, we required eroding grains
%  that were sufficiently separated to guarantee that the trapezoid rule
%  provides a sufficient amount of accuracy. We now remove this
%  restriction by introducing a Barycentric quadrature rule whose error
%  is bounded independent of the geometry.  The Barycentric quadrature
%  rule is computationally expensive, so it is only used when necessary,
%  and the spectrally accurate fast-multipole-accelerated trapezoid rule
%  is used otherwise. We also simulate and analyze transport through
%  eroded porous media.  By using the Barycentric quadrature rule, the
%  analysis examines flow both in the fast and slow regions with
%  high-order accuracy. Our results reveal several interesting  effects
%  of erosion on the tortuosity, dispersion, and pore size distributions.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\citep{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\citep{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude.  Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\citep{saf1959}, (ii) quantify
mixing~\citep{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) develop meaningful constitutive relationships that link the
microscopic and macroscopic realms~\citep{mil-chr-imh-mcb-ped1998}.
Examples of coarse-grained models for porous media flow include
permeability-porosity relationships~\citep{dar-mcc1998, car1937},
tortuosities~\citep{mat-kha-koz2008, dud-koz-mat2011, kop-kat-tim1996},
geometry connectivity~\citep{knu-car2005}, anomalous
dispersion~\citep{den-cor-sch-ber2004}, and more.

@@

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Contributions}
%Our first contribution is extending the Barycentric quadrature rule
%of~\citet{bar-wu-vee2015} to accurately compute velocity gradients
%associated with Stokes flow. This new quadrature rule permits accurate
%and stable simulations of erosion, and, moreover, enables visualization
%of the vorticity field throughout.
%
%Our second contribution is computing and analyzing trajectories in
%eroded geometries.  These trajectories are formed by combining the
%Barycentric quadrature rule and a fourth-order Runge-Kutta method. Then,
%they are used to compute tortuosities and dispersion rates of eroded
%geometries. Finally, we examine the effect of erosion on the pore size
%distribution.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Limitations}
%As in our previous work~\citep{qua-moo2018}, we consider two-dimensional
%immobile eroding grains. However, this work does not limit the proximity
%of  grains. Compared to our previous work, we are able to simulate much
%denser suspensions with a comparable number of discretization points and
%without upsampling.  While we do investigate transport in this work, we
%only consider passive particles that are advected by the flow.  To model
%other physical processes, such as contaminant transport, we plan to
%simulate a finite Peclet number advection-diffusion equation in eroded
%geometries.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\subsection{Outline of the Paper}
This paper is organized as follows. In section~\ref{sec:formulation}, we
summarize the erosion model that is described in more detail in previous
work~\citep{qua-moo2018}.  In section~\ref{sec:DLP}, we recast all the
governing equations as layer potentials defined in both $\RR^2$ and in
$\CC$.  Section~\ref{sec:transport} describes measures for
characterizing the geometry and transport. Section~\ref{sec:method}
describes the numerical methods, with special attention paid to the new
quadrature method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  All eroding
bodies are placed in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the
more complex flow intervening between the bodies. Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, and $\Vn$ is the normal
velocity of $\gamma$. The shear stress on $\gamma$ is
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
where, $\nn$ is the normal vector pointing into the body, and $\ss$ is
the unit tangent vector pointing in the counterclockwise direction. We
simulate erosion by alternating between solving the fluid equations and
advancing the eroding grains.  The strength of $\UU$ is adjusted at each
time step to achieve a constant pressure drop across the channel,
motivated by the geological situation of a porous medium connecting two
regions of  fixed hydraulic heads.

@@

To establish an asymptotic anomalous dispersion rate, the trajectories
must pass several grains. The geometries that we consider are too short
to observe asymptotic dispersion, so we use a reinsertion method to form
longer trajectories. Similar to the work of
others~\citep{dea-qua-bir-jua2018, puy-gou-den2019}, once a particle
reaches the outlet of the porous region, it is reinserted at the inlet.
To minimize errors caused by reinsertion, the particle is initialized at
the discretization point that has the closest velocity to the particle's
velocity at the outlet.  After a single trajectory is formed, it has
undergone a collection of reinsertions.  Then, as a post-processing
step, the trajectory is made continuous by attaching the tail of each
segment to the origin of the next segment.

%We consider trajectories in an open channel so that they can be compared
%to trajectories in eroded geometries with large permeability.  The
%trajectories are $(s_1(t),s_2(t)) = ((1-s_2(0)^2)t + s_1(0), s_2(0))$.
%%\begin{align}
%%  s_1(t) = (1-s_2(0)^2)t + s_1(0), \quad
%%  s_2(t) = s_2(0).
%%\end{align}
%Averaging over $s_2(0) \in [-1,1]$, the first and second ensemble
%moments are
%\begin{align}
%  \langle \lambda \rangle (t) = \frac{2}{3}t, \quad 
%    \sigma_\lambda^2(t) = \frac{4}{45}t^2.
%\end{align}
%In section~\ref{sec:results}, we compare the second moment in an open
%channel with the trajectories in an eroded geometry.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\citep{dud-koz-mat2011}.  The
hydraulic tortuosity is defined by taking the average over all points on
the inlet cross-section
\begin{align}
  T = \frac{1}{d}\left(\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0 \right)
  \Bigg/
  \left(\int_{S}u_1(x_0,y_0)\,dy_0 \right),
%  T = \frac{1}{d}\frac{\displaystyle\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0}
%  {\displaystyle\int_{S}u_1(x_0,y_0)\,dy_0},
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.

The tortuosity can also be computed with an area integral. Assuming that
the flow is incompressible and not re-entrant, meaning that all
streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent
to~\citep{dud-koz-mat2011}
\begin{align}
  T = \left(\int_\Omega \|\uu(\yy)\|\, d\yy \right)
    \Bigg/
      \left(\int_\Omega u_1(\yy)\, d\yy \right),
%  T = \frac{\displaystyle\int_\Omega \|\uu(\yy)\|\, d\yy}
%           {\displaystyle\int_\Omega u_1(\yy)\, d\yy},
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\citep{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in section~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes between
neighboring grains. Two grains are neighbors if they share an edge of
the Delanuay triangulation with nodes at the center of each eroding
grain. (b) The pore sizes of an eroded geometry. The black curve is the
Weibull distribution with the same first two moments as the data.}
\end{figure}
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. The pores of an eroded
geometry are illustrated in figure~\ref{fig:Eroding100gap}(a). We do not
consider pores between eroding bodies and the solid wall $\Gamma$, so
some of the grains near the porous region boundary only have two
neighbors. Having defined the pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}.  In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.

@@

Again, given a complex-valued density function $\eta$, the boundary data
of $v[\eta](x)$, as defined in equation~\eqref{eqn:laplaceComplex},
satisfies the Sokhotski-Plemelj jump relation~\eqref{eqn:SPrelation}.
Since the limiting boundary data of $v$ differs when considering $x \in
\Omega$ and $x \in \Omega^c$, we denote the boundary data as $v^-$ for
$x \in \Omega$, and as $v^+$ for $x \in \Omega^c$.  Rather than directly
applying the trapezoid rule to approximate $v(x)$ in
equation~\eqref{eqn:cauchy}, we start with the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^{-}(y) - v(x)}{y - x} \,dy = 0,
    \quad x \in \Omega.
\end{align}
Since the integrand is bounded for all $x \in \Omega$, we can apply the
trapezoid rule
\begin{align}
  \sum_{j=1}^{N} \frac{v^{-}(y_j) - v(x)}{y_j - x} w_j \approx 0,
\end{align}
and the error is independent of $x$.  Rearranging for $v(x)$, we have
the interior Barycentric quadrature rule
\begin{align}
  v(x) = \left(\sum_{j=1}^N \frac{v^{-}(y_j)}{y_j - x} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^N \frac{1}{y_j - x} w_j \right), 
  \quad x \in \Omega.
%  v(x) = \frac{\sum\limits_{j=1}^N \frac{v^{-}(y_j)}{y_j - x} w_j}
%  {\sum\limits_{j=1}^N \frac{1}{y_j - x} w_j}, \quad x \in \Omega.
  \label{eqn:BaryvInterior}
\end{align}
Using a similar construction and letting $a$ be any point inside
$\Omega$, the exterior Barycentric quadrature rule is
\begin{align}
  v(x) = \frac{1}{x-a} 
    \left(\sum_{j=1}^N \frac{v^+(y_j)}{y_j - x}w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j - a)^{-1}}{y_j - x}w_j \right),
    \quad x \in \Omega^c.
%  v(x) = \frac{1}{x-a} 
%    \frac{\sum\limits_{j=1}^N \frac{v^+(y_j)}{y_j - x}w_j}
%    {\sum\limits_{j=1}^N \frac{(y_j - a)^{-1}}{y_j - x}w_j},
%    \quad x \in \Omega^c.
  \label{eqn:BaryvExterior}
\end{align}

>>>>

\documentclass{jfm}

@@

@@

\begin{abstract} 
  Transport of viscous fluid through porous media is a direct
  consequence of the pore structure. Here we investigate transport
  through a specific class of two-dimensional porous geometries, namely
  those formed by fluid-mechanical erosion.  We investigate the
  tortuosity and dispersion by analyzing the first two statistical
  moments of tracer trajectories. For most initial configurations,
  tortuosity decreases in time as a result of erosion increasing the
  porosity.  However, we find that tortuosity can also increase
  transiently in certain cases.  The porosity-tortuosity relationships
  that result from our simulations are compared with models available in
  the literature.  Asymptotic dispersion rates are also strongly
  affected by the erosion process, as well as by the number and
  distribution of the eroding bodies. Finally, we analyze the pore size
  distribution of an eroding geometry. The simulations are performed by
  combining a high-fidelity boundary integral equation solver for the
  fluid equations, a second-order stable time stepping method to
  simulate erosion, and new numerical methods to stably and accurately
  resolve nearly-touching eroded bodies and particle trajectories near
  the eroding bodies.
\end{abstract}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}
\label{sec:intro}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Porous media flow plays an important role in many environmental and
industrial applications.  Depending on the application, length scales
can vary from $10^{-6}\mathrm{m}$ to
$10^{-1}\mathrm{m}$~\citep{mil-chr-imh-mcb-ped1998} and velocity scales
can be as small as $10^{-1}\mathrm{m/day}$~\citep{kut-scr-dav-ham1995}.
Moreover, for a single porous geometry, the pore sizes and velocities
can range over several orders of magnitude. Numerical methods that
resolve this range of scales offer the ability to: (i) characterize
dispersion~\citep{saf1959}, (ii) quantify
mixing~\citep{leb-den-dav-bol-car-dec-bou2011, den-leb-eng-bij2011}, and
(iii) develop meaningful constitutive relationships that link the
microscopic and macroscopic realms~\citep{mil-chr-imh-mcb-ped1998}.
Examples of coarse-grained models for porous media flow include
tortuosities~\citep{mat-kha-koz2008, dud-koz-mat2011, kop-kat-tim1996},
geometry connectivity~\citep{knu-car2005}, anomalous
dispersion~\citep{den-cor-sch-ber2004}, and permeability-porosity
relationships~\citep{dar-mcc1998, car1937}, where the porosity is the
fraction of the geometry that is fluidized.

@@

While natural porous media are three-dimensional, there are several
challenges that must be overcome to extend our high-order methods to
three dimensions.  A three-dimensional version of a {\thL} formulation
exists~\citep{amb-sie-tlu2013}, but the Barycentric quadrature method
relies upon a complex variable formulation of the BIE, making it
inherently two-dimensional. Therefore, a different quadrature method is
required. Additional challenges of three-dimensional simulations include
spectral representations of the geometry and efficiently solving
discretizations of the BIE. While high-order methods for
three-dimensional BIEs have been developed, their computational cost
limits the number of bodies that can be practically simulated. Another
limitation of this work is the assumption of immobile grains.
Mechanisms such as sedimentation play a role in real applications, and
this will effect the transport properties. Simulating mobile grains
introduces the additional challenge of resolving contact.  Since
high-order methods have never been used to study transport in eroded
porous media, as a first step we consider the case  with two-dimensional
immobile grains. This assumption allows for a clear interpretation of
the results.  For example, in this work we are able to simulate a large
densely-packed suspension (up to $100$ in this manuscript).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
This paper is organized as follows. In section~\ref{sec:formulation}, we
summarize the erosion model that is described in more detail in previous
work~\citep{qua-moo2018}.  In section~\ref{sec:DLP}, we recast all the
governing equations as layer potentials defined in both $\RR^2$ and in
$\CC$.  Section~\ref{sec:transport} describes measures for
characterizing the geometry and transport. Section~\ref{sec:method}
describes the numerical methods, with special attention paid to the new
quadrature method for computing the shear stress and the vorticity.
Section~\ref{sec:results} presents numerical examples for a variety of
dense packings of bodies.  Finally, concluding remarks are made in
section~\ref{sec:conclusions}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Governing Equations}
\label{sec:formulation}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
We start by defining the main variables used to model erosion.  We only
briefly summarize the model, and a more detailed description can be
found in previous work~\citep{qua-moo2018}.  We consider flows inside a
confined geometry $\Omega$ that contains $M$ eroding bodies with
boundaries $\gamma_\ell$, $\ell = 1,\ldots,M$.  The boundary of the
fluid domain is $\bd \Omega = \Gamma \cup \gamma_1 \cup \cdots \cup
\gamma_M$, where $\Gamma$ is the outer boundary, taken to be a slightly
smoothed version of the boundary of $[-3,3] \times [-1,1]$.  All eroding
bodies are placed in $[-1,1] \times [-1,1]$ to create a buffer region
that allows the flow profile imposed at the inlet to transition to the
more complex flow intervening between the bodies. Neglecting inertial
forces, the governing equations are
\begin{equation}
\label{eqn:erosionModel}
  \begin{split}
    \mu \Delta \uu = \grad p, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of momentum}, \\
    \grad \cdot \uu = 0, &\hspace{20pt} \xx \in \Omega, \gap 
      &&\mbox{\em conservation of mass}, \\
    \uu = \mathbf{0}, &\hspace{20pt} \xx \in \gamma, \gap 
      &&\mbox{\em no slip on the eroding bodies}, \\
    \uu = \UU, &\hspace{20pt} \xx \in \Gamma, \gap 
      &&\mbox{\em outer wall velocity}, \\
    \Vn = \CE \, \abs{\tau}, &\hspace{20pt} \xx \in \gamma,
      &&\mbox{\em erosion model}.
  \end{split}
\end{equation}
Here $\uu$ is the fluid velocity, $p$ is the pressure, $\UU$ is a
prescribed Hagen-Poiseuille velocity field, and $\Vn$ is the normal
velocity of $\gamma$. Since the rate of erosion is much slower than the
background flow, a quasi-steady approximation justifies the no-slip
boundary condition. Parameters include the fluid viscosity $\mu$ and the
material-dependent erosion constant $\CE$. By non-dimensionalizing
equations~\eqref{eqn:erosionModel}~\citep{qua-moo2018}, we set both
these parameters to one. The shear stress on $\gamma$ is
\begin{align}
  \tau = -(\nabla \uu + \nabla \uu^T) \nn \cdot \ss
  \label{eqn:shearStress}
\end{align}
where, $\nn$ is the normal vector pointing into the body, and $\ss$ is
the unit tangent vector pointing in the counterclockwise direction. We
simulate erosion by alternating between solving the fluid equations and
advancing the eroding grains.  The strength of $\UU$ is adjusted at each
time step to achieve a constant pressure drop across the channel,
motivated by the geological situation of a porous medium connecting two
regions of  fixed hydraulic heads.

@@

To establish an asymptotic anomalous dispersion rate, the trajectories
must pass several grains. The geometries that we consider are too short
to observe asymptotic dispersion, so we use a reinsertion method to form
longer trajectories. Similar to the work of
others~\citep{dea-qua-bir-jua2018, puy-gou-den2019}, once a particle
reaches the outlet of the porous region, it is reinserted at the inlet.
To minimize errors caused by reinsertion, the particle is initialized at
the discretization point that has the closest velocity to the particle's
velocity at the outlet. This method of reinsertion does not artificially
decorrelate the particle velocities~\citep{puy-gou-den2019}, and this
allows us to reliably use the statistical moments~\eqref{eqn:moments} to
characterize the anomalous dispersion. After a single trajectory is
formed, it has undergone a collection of reinsertions.  Then, as a
post-processing step, the trajectory is made continuous by attaching the
tail of each segment to the origin of the next segment.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Tortuosity}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
The tortuosity is a dimensionless number that quantifies the amount of
twisting of streamlines. Unlike the dispersion calculations, we do not
use reinsertion to form long trajectories.  The local tortuosity is
\begin{align}
  \tau(y_0) = \frac{\lambda(y_0)}{d}.
  \label{eqn:localTort}
\end{align}
Here the streamline originates on the inlet cross-section $x=x_0$ at
$(x_0,y_0)$, and its arclength, $\lambda(y_0)$, is calculated until the
streamline passes the parallel outlet cross-section $x = x_0 + d$.  In
this work, we consider streamlines originating at $x=-1$ and terminating
at $x=1$, so $d=2$.  However, other choices for the terminal point when
computing the tortuosity are sometime used~\citep{dud-koz-mat2011}.
We note that if a streamline originates in a recirculation zone, which
is possible in two dimensions~\citep{hig1985}, then the local tortuosity
will be infinite. However, in each of our numerical experiments, all the
streamlines connect the inlet to the outlet. Having only finite local
tortuosities, the hydraulic tortuosity is defined by taking the average
over all points on the inlet cross-section
\begin{align}
  T = \frac{1}{d}\left(\int_{S}u_1(x_0,y_0)\lambda(y_0)\,dy_0 \right)
  \Bigg/
  \left(\int_{S}u_1(x_0,y_0)\,dy_0 \right),
  \label{eqn:tortuosity1}
\end{align} 
where $S$ is the inlet cross-section $x = -1$, $u_1(y_0)$ is the
$x$-component of the velocity at the initial point of the streamline,
and $d=2$ is the distance between the inlet and outlet.  Note that $T
\geq 1$, and $T=1$ only if no grains are present.

The tortuosity can also be computed with an area integral. Assuming that
the flow is incompressible and not re-entrant, meaning that all
streamlines connect the two cross-sections, the tortuosity in
equation~\eqref{eqn:tortuosity1} is equivalent
to~\citep{dud-koz-mat2011}
\begin{align}
  T = \left(\int_\Omega \|\uu(\yy)\|\, d\yy \right)
    \Bigg/
      \left(\int_\Omega u_1(\yy)\, d\yy \right),
  \label{eqn:tortuosity2}
\end{align}
where $\Omega$ is the fluid region between the inlet and outlet
cross-sections.  Recirculation zones are possible in viscous
fluids~\citep{hig1985}, but they are very small in the examples we
consider and have a negligible effect on the tortuosity.  Since
equation~\eqref{eqn:tortuosity2} does not require the additional work of
computing particle trajectories at every time step, we use this
definition for the majority of the tortuosity calculations.  However, we
do compare the two definitions for the tortuosity at several porosities
in section~\ref{sec:results}.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Pore Throat Size}
\label{sec:throats}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height =0.8\linewidth]{./figs/triangulation_100b100}
\caption{}
\end{subfigure}
\begin{subfigure}[b]{0.5\textwidth}
\includegraphics*[height = 0.8\linewidth]{./figs/gap_hist100b100}
\caption{}
\end{subfigure}
\caption{\label{fig:Eroding100gap} (a) The pore sizes, represented by
the length of the red line segments, between neighboring grains. Two
grains are neighbors if they share an edge of the Delanuay triangulation
with nodes at the center of each eroding grain. (b) The pore sizes of an
eroded geometry. The black curve is the Weibull distribution with the
same first two moments as the data.}
\end{figure}
While transport in porous media depends on the porosity, it also depends
on the placement of the grains.  In particular, grain placement affects
velocity scales~\citep{ali-par-wei-bre2017}, correlation
structures~\citep{leb-ded-dav-bou2007}, contaminant
transport~\citep{knu-car2005},
channelization~\citep{sie-ili-pri-riv-gua2019,berhanu2012shape}, and
pore network models~\citep{bry-kin-mel1993, bry-mel-cad1993,
bij-blu2006}. To characterize the grain placement, we compute
distributions of pore sizes between neighboring grains.  To define
neighboring grains, we form the Delaunay triangulation using nodes
placed at the center of each eroding grain and at a collection of points
around the boundary of the porous media.  Then, we say that two grains
are neighbors if their centers share an edge of the
triangulation~\citep{dea-qua-bir-jua2018}. Finally, we define the pore
sizes to be the minimum distance between all pairs of neighboring
grains, and these distances are represented by the red line segments in
figure~\ref{fig:Eroding100gap}(a). We do not consider pores between
eroding bodies and the solid wall $\Gamma$, so some of the grains near
the porous region boundary only have two neighbors. Having defined the
pore sizes, we plot its distribution in
figure~\ref{fig:Eroding100gap}(b) and compare it with the Weibull
distribution, a distribution used by others to characterize pore
sizes~\citep{ioa-cha1993}. In section~\ref{sec:Eroding100}, we
investigate the effect of erosion on the pore size distribution.

@@

Again, given a complex-valued density function $\eta$, the boundary data
of $v[\eta](x)$, as defined in equation~\eqref{eqn:laplaceComplex},
satisfies the Sokhotski-Plemelj jump relation~\eqref{eqn:SPrelation}.
Since the limiting boundary data of $v$ differs when considering $x \in
\Omega$ and $x \in \Omega^c$, we denote the boundary data as $v^-$ for
$x \in \Omega$, and as $v^+$ for $x \in \Omega^c$.  Rather than directly
applying the trapezoid rule to approximate $v(x)$ in
equation~\eqref{eqn:cauchy}, we start with the identity
\begin{align}
  \int_{\bd\Omega} \frac{v^{-}(y) - v(x)}{y - x} \,dy = 0,
    \quad x \in \Omega.
\end{align}
Since the integrand is bounded for all $x \in \Omega$, we can apply the
trapezoid rule
\begin{align}
  \sum_{j=1}^{N} \frac{v^{-}(y_j) - v(x)}{y_j - x} w_j \approx 0,
\end{align}
and the error is independent of $x$.  Rearranging for $v(x)$, we have
the interior Barycentric quadrature rule
\begin{align}
  v(x) = \left(\sum_{j=1}^N \frac{v^{-}(y_j)}{y_j - x} w_j \right)
  \Bigg/
  \left(\sum_{j=1}^N \frac{1}{y_j - x} w_j \right), 
  \quad x \in \Omega.
  \label{eqn:BaryvInterior}
\end{align}
Using a similar construction and letting $a$ be any point inside
$\Omega$, the exterior Barycentric quadrature rule is
\begin{align}
  v(x) = \frac{1}{x-a} 
    \left(\sum_{j=1}^N \frac{v^+(y_j)}{y_j - x}w_j \right)
    \Bigg/
    \left(\sum_{j=1}^N \frac{(y_j - a)^{-1}}{y_j - x}w_j \right),
    \quad x \in \Omega^c.
  \label{eqn:BaryvExterior}
\end{align}
